---
title: "GIPR Liraglutide analysis"
author: "Erik Ländström"
date: "7 March 2019"
output: 
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries 

```{r library, message = FALSE}
library(tidyverse)
library(biomaRt)
library(knitr)
library(kableExtra)
library(DEP)
library(readxl)
library(qvalue)
library(kableExtra)
library(naniar)
```

# 1 Preprocessing
Read data with gene names and GOs from biodbnet["https://biodbnet-abcc.ncifcrf.gov/db/db2db.php"].

```{r read_data}
# Read data
data_raw <- read_tsv("proteinGroups_with_gene_names_and_GOs.txt")

# Quick look at data columns
glimpse(data_raw)
```

Remove all protein groups without a gene name (or locus name).

```{r remove}
filtered_data <- data_raw %>% 
  dplyr::filter(!is.na(gene_name))
```

Get gene names from biomaRt and replace any locus names with the new gene names.

```{r}
# Remove isoform from refseq identifier
filtered_data <- filtered_data %>%
  mutate(Protein_ID = str_extract(`Majority protein IDs`, "NP_[0-9]+|XP_[0-9]+"))
```

Extract gene names from biomart.

```{r mart_func, echo = FALSE}
convert_NCBIIDs_to_gene_names <- function(tb) {
  
  # Load sus scrofa dataset from biomaRt and change column name to Protein_ID
  ensembl <- useMart("ensembl", dataset = "sscrofa_gene_ensembl")  
  
  # Get gene names for peptide IDs
  NP <- as_tibble(getBM(attributes = c("refseq_peptide", "external_gene_name"),
              filters = "refseq_peptide",
              values = tb,
              mart = ensembl
              ))
  
  NP <- NP %>%
    rename(Protein_ID = refseq_peptide)
  
  # Load sus scrofa dataset from biomaRt and change column name to Protein_ID
  XP <- as_tibble(getBM(attributes = c("refseq_peptide_predicted", "external_gene_name"),
              filters = "refseq_peptide_predicted",
              values = tb,
              mart = ensembl
              ))
  
  XP <- XP %>%
    rename(Protein_ID = refseq_peptide_predicted)
  
  # Use full_join to join the two tables
  gene_names <- full_join(NP, XP, by = "Protein_ID")
  
  # Tidy data, gene names in one column and remove NAs
  gene_names <- gene_names %>%
    gather(external_gene_name.x, external_gene_name.y, key = "Source", value = "Gene_name") %>%
    dplyr::select(Protein_ID, Gene_name) %>%
    dplyr::filter(!is.na(Gene_name)) 
  
  # If there are any "" artifacts from the biomaRt search, replace them with NA
  gene_names[gene_names == ""] <- NA
  
  # Save output in a tibble
  gene_names <<- gene_names
}
```

```{r mart_convert}
# Find gene names using biomart
convert_NCBIIDs_to_gene_names(filtered_data$Protein_ID)

# Join with filtered data
filtered_data <- full_join(filtered_data, gene_names, by = "Protein_ID")
```

Replace loci and months with gene names from biomaRt (if any).

```{r sub_biomart_names}
data <- filtered_data %>% 
  rename(gene_name_biomart = Gene_name) %>%
  mutate(gene = if_else(str_detect(gene_name, "LOC.+") & !is.na(gene_name_biomart), gene_name_biomart, gene_name),
         gene = if_else(str_detect(gene, "[A-Za-z]+ [0-9]+"), gene_name_biomart, gene))
```

```{r unique}
# Make unique gene names (not sure if this is necessary with DEP)
data <- make_unique(data, "gene", "Protein_ID")
```

## Create experimental design

Create an experimental design file for `DEP`.

```{r read_animals, eval = FALSE}
library(readxl)

animal_list <- read_xlsx("Animal List young group.xlsx")

animal_list %>% 
  mutate(Group = str_extract(Group, "[A-Z]"),
         Gender = str_extract(Gender, "[a-z]"),
         Litter = str_extract(Litter, "[0-9]")) %>% 
  dplyr::filter(ID != 1495) %>% # Since it was a muscle sample
  unite(col = "label", Group, ID, sep = "_", remove = FALSE) %>% 
  unite(col = "group", Group, Gender, Series, Litter) %>% 
  rename(replicate = ID,
         condition = group) %>% 
  dplyr::select(label, condition, replicate) %>% 
  write_tsv("experimental_design_GIPR_liraglutide.txt")
```

Create a summarized experiment.

```{r SExperiment}
# Grep for lfq columns
lfq_columns <- grep("LFQ.", colnames(data))

# Read experimental design
experimental_design <- read_tsv("experimental_design_GIPR_liraglutide.txt")
experimental_design

# Generate summarized experiment
data_se <- make_se(data, lfq_columns, experimental_design)
data_se
```

# 2 Analysis of summarized experiment

## 2.1 Missing value filtering

Plot protein identification overlap plot.

```{r frequency}
plot_frequency(data_se)
```

Filter data on valid values.

```{r count_data}
# Calculate valid values per protein and treatment 
count_data <- data %>% 
  gather(`LFQ intensity L_1312`:`LFQ intensity P_1505`, key = "sample", value = "intensity") %>% 
  dplyr::select(name, sample, intensity) %>% 
  mutate(treatment = str_extract(sample, "[A-Z]_"),
         intensity = log2(intensity)) %>% 
  replace_with_na(replace = list(intensity = -Inf)) %>% 
  group_by(name, treatment) %>% 
  summarise(counts = sum(!is.na(intensity)))
```

I will choose valid values 5 per group, that was what I did last time in 
Perseus, and 1 missing value to compare with.

```{r count_vv}
# Filter for 5 valid values
list_5vv <- count_data %>% 
  spread(key = "treatment", value = "counts") %>% 
  dplyr::filter(L_ >= 5 | P_ >= 5)

# Filter for one missing value
list_1mv <- count_data %>% 
  spread(key = "treatment", value = "counts") %>% 
  dplyr::filter(L_ >= 7 | P_ >= 8)
```

## 2.2 Analysis of 5 valid values

```{r make_se}
data_5vv <- left_join(list_5vv, data, by = "name")

# Grep for lfq columns
lfq_columns <- grep("LFQ.", colnames(data_5vv))

data_5vv_se <- make_se(data_5vv, lfq_columns, experimental_design)
```

```{r plots 2}
plot_frequency(data_5vv_se)
plot_numbers(data_5vv_se)
plot_coverage(data_5vv_se)
plot_detect(data_5vv_se)
```

Normalize with `vsn`.

```{r norm}
data_5vv_norm <- normalize_vsn(data_5vv_se)
meanSdPlot(data_5vv_se)
meanSdPlot(data_5vv_norm)
plot_normalization(data_5vv_se, data_5vv_norm)
```

```{r plot_mv2}
plot_missval(data_5vv_norm)
```

Imputation.

```{r}
imputed_data_5vv <- impute(data_5vv_norm, fun = "man", shift = 1.8, scale = 0.3)

plot_imputation(data_5vv_norm, imputed_data_5vv)
```

Save imputed data.

```{r}
imputed_data_5vv_long <- get_df_long(imputed_data_5vv)

write_tsv(imputed_data_5vv_long, "GIPR_liraglutide_liver_5vv_imputed.tsv")
```

## 2.2 Analysis of 5 valid values

```{r make_se1}
data_1mv <- left_join(list_1mv, data, by = "name")

# Grep for lfq columns
lfq_columns <- grep("LFQ.", colnames(data_1mv))

data_1mv_se <- make_se(data_1mv, lfq_columns, experimental_design)
```

```{r plots 1}
plot_frequency(data_1mv_se)
plot_numbers(data_1mv_se)
plot_coverage(data_1mv_se)
plot_detect(data_1mv_se)
```

Normalize with `vsn`.

```{r norm2}
data_1mv_norm <- normalize_vsn(data_1mv_se)
meanSdPlot(data_1mv_se)
meanSdPlot(data_1mv_norm)
plot_normalization(data_1mv_se, data_1mv_norm)
```

```{r plot_mv}
plot_missval(data_1mv_norm)
```

Imputation.

```{r}
imputed_data_1mv <- impute(data_1mv_norm, fun = "man", shift = 1.8, scale = 0.3)

plot_imputation(data_1mv_norm, imputed_data_1mv)
```

Save imputed data.

```{r}
imputed_data_1mv_long <- get_df_long(imputed_data_1mv)

write_tsv(imputed_data_1mv_long, "GIPR_liraglutide_liver_1mv_imputed.tsv")
```




























