---
title: "GIPR statistical analysis 5vv"
author: "Erik Ländström"
date: "8 March 2019"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Read libraries.

```{r library, message = FALSE}
library(tidyverse)
library(modelr)
library(broom)
library(ggfortify)
library(sva)
library(qvalue)
```

Read functions.

```{r functions, echo = FALSE}
source("functions/separate_condition_from_DEP_output.R")
source("functions/format_data_for_PCA.R")
source("functions/plot_var_exp.R")
source("functions/make_var_exp_kable.R")
source("functions/calculate_variance_explained.R")
source("functions/plot_pca.R")
source("functions/plot_multiple_PCAs.R")
source("functions/multiple_ttests.R")
source("functions/fdr_correction_with_qvalue.R")
source("functions/plot_pvalue_histogram.R")
```

# Analysis

Perform statistical analysis of imputed data with at least 5 valid values per 
group.

Read data.

```{r message = FALSE}
data <- read_tsv("GIPR_liraglutide_liver_5vv_imputed.tsv")
```

Separate conditions into separate columns.

```{r sep_col}
factored_data <- separate_condition_from_DEP_output(data, c("treatment", "gender", "series", "litter"), remove = FALSE)
```


## Linear model

Nest data.

```{r}
nested_data <- factored_data %>% 
  group_by(name) %>% 
  nest()
```

Linear model function.

```{r func}
protein_model <- function(tb) {
  lm(LFQ ~ gender * litter, data = tb)
}
```

Perform linear regression on each protein and calculate/extract residuals.

```{r lming}
nested_model_data <- nested_data %>% 
  mutate(model = map(data, protein_model))

nested_resids_data <- nested_model_data %>% 
  mutate(resids = map2(data, model, add_residuals))

resids <- unnest(nested_resids_data, resids)
```

Calculate variance and compare with original data.

```{r var}
variance <- resids %>% 
  group_by(name) %>% 
  summarise(var_lfq = var(LFQ),
            var_resid = var(resid))

variance %>% 
  summarise(mean_lfq = mean(var_lfq),
            mean_resid = mean(var_resid))

# Test if the regression reduced the variance for all proteins
variance %>% 
  mutate(smaller = if_else(var_resid < var_lfq, 1, 0)) %>% 
  summarise(n = sum(smaller),
            all = n())

resids %>% 
  group_by(name, treatment) %>% 
  summarise(var_lfq = var(LFQ),
            var_resid = var(resid))
```

The linear model decreased the variance for every protein!

## PCA

Unite replicate (ID) with condition (grouping), and spread to intensities into
separate columns for each sample.

```{r pca_prep}
lfq_for_pca <- resids %>% 
  dplyr::select(name, replicate, condition, LFQ) %>% 
  unite(replicate, replicate, condition) %>% 
  spread(key = "replicate", value = "LFQ")

resids_for_pca <- resids %>% 
  dplyr::select(name, replicate, condition, resid) %>% 
  unite(replicate, replicate, condition) %>% 
  spread(key = "replicate", value = "resid")
```


Perform PCA for both LFQ and residuals.

```{r pca}
lfq_pca <- format_data_for_PCA(lfq_for_pca, 17, c("sample", "treatment", "gender", "series", "litter"))

resid_pca <- format_data_for_PCA(resids_for_pca, 17, c("sample", "treatment", "gender", "series", "litter"))
```


Calculate the variance explained by each principal component for original and
data processed with linear regression.

```{r var_exp}
# Calculate the variance that each PC explains
lfq_var_exp <- calculate_variance_explained(lfq_pca)
make_var_exp_kable(lfq_var_exp)

resid_var_exp <- calculate_variance_explained(resid_pca)
make_var_exp_kable(resid_var_exp)

# Plot variance explained plots
plot_var_exp(lfq_var_exp)
plot_var_exp(resid_var_exp)
```

The principal components from the linear model explains considerably more of the
variance than the LFQ intensities.


Plot the first two PCs of the PCA.

```{r plot_pca}
plot_pca(lfq_pca, "LFQ", "treatment")
plot_pca(resid_pca, "resids", "treatment")
```



Plot PC1 to PC4.

```{r plot_pcas}
plot_multiple_PCAs(lfq_pca, 1, 4, "treatment")
plot_multiple_PCAs(resid_pca, 1, 4, "treatment")
```


# Surrogate variable analysis

Format data for sva.

Create expression data.

```{r exp_data}
exp_data <- resids %>% 
  dplyr::select(name, replicate, LFQ) %>% 
  spread(key = "replicate", value = "LFQ")

row.names(exp_data) <- exp_data$name

exp_data <- exp_data %>% 
  dplyr::select(-name)
```

Create phenotypic data.

```{r pheno}
pheno_data <- resids %>% 
  dplyr::select(treatment, gender, series, litter, replicate) %>% 
  distinct() %>% 
  mutate(sample = row_number())

rownames(pheno_data) <- pheno_data$replicate

pheno_data <- pheno_data %>% 
  dplyr::select(sample, treatment, gender, series, litter)
```

Create linear models, full and null model.

```{r lm}
mod = model.matrix(~ as.factor(treatment), data = pheno_data)

mod0 = model.matrix(~ 1,data = pheno_data)
```

Sva using LFQ values

```{r sva}
svobj = sva(as.matrix(exp_data), mod, mod0, n.sv= 2)
svobj[[1]]
```

Sva using LFQ plus the known batch effects: gender and series.

```{r lm_known}
mod = model.matrix(~ as.factor(treatment) * as.factor(gender) * as.factor(series), data = pheno_data)

mod0 = model.matrix(~ as.factor(gender) * as.factor(series),data = pheno_data)

svobj = sva(as.matrix(exp_data), mod, mod0, n.sv= 2)
svobj[[1]]
```

Sva analysis using residuals from linear model.

```{r sva_lm}
exp_resids <- resids %>% 
  dplyr::select(name, replicate, resid) %>% 
  spread(key = "replicate", value = "resid")

row.names(exp_resids) <- exp_resids$name

exp_resids <- exp_resids %>% 
  dplyr::select(-name)

mod = model.matrix(~ as.factor(treatment), data = pheno_data)

mod0 = model.matrix(~ 1,data = pheno_data)

svobj = sva(as.matrix(exp_resids), mod, mod0, n.sv= 2)
svobj[[1]]

sv_factor1 <- tibble(replicate = as.numeric(rownames(pheno_data)),
       sv1 = svobj[[1]][, 1])
```

Join sva factor sv1 with previous results.

```{r sva_resid}
resids_sv <- full_join(resids, sv_factor1, by = "replicate")
```

Nest data.

```{r nest}
nested_data <- resids_sv %>% 
  group_by(name) %>% 
  rename(resid_old = resid) %>% 
  nest()
```

Linear model including the sv1 factor.

```{r func2}
protein_model <- function(tb) {
  lm(resid_old ~ sv1, data = tb)
}
```

Perform linear model.

```{r lming2}
nested_model_data <- nested_data %>% 
  mutate(model = map(data, protein_model))

nested_resids_data <- nested_model_data %>% 
  mutate(resids = map2(data, model, add_residuals))

residuals_sv <- unnest(nested_resids_data, resids)
```

Calculate variance.

```{r var2}
variance <- resids %>% 
  group_by(name) %>% 
  summarise(var_lfq = var(LFQ),
            var_resid = var(resid))

variance %>% 
  summarise(mean_var_lfq = mean(var_lfq),
            mean_var_resid = mean(var_resid))

variance %>% 
  mutate(smaller = if_else(var_resid < var_lfq, 1, 0)) %>% 
  summarise(n = sum(smaller))

resids %>% 
  group_by(name, treatment) %>% 
  summarise(var_lfq = var(LFQ),
            var_resid = var(resid))
```

Calculate variance explained.

```{r var_exp2}
resids_sv_for_pca <- residuals_sv %>% 
  dplyr::select(name, replicate, condition, resid) %>% 
  unite(replicate, replicate, condition) %>% 
  spread(key = "replicate", value = "resid")

resid_sv_pca <- format_data_for_PCA(resids_sv_for_pca, 17, c("sample", "treatment", "gender", "series", "litter"))

resid_sv_var_exp <- calculate_variance_explained(resid_sv_pca)
make_var_exp_kable(resid_sv_var_exp)


plot_var_exp(resid_sv_var_exp)
```

```{r}
plot_pca(resid_sv_pca, "resids", "treatment")

plot_multiple_PCAs(resid_pca, 1, 4, "treatment")
plot_multiple_PCAs(resid_sv_pca, 1, 4, "treatment")
```

# t-test

Perform multiple t-test.

```{r ttest}
residuals_sv_ttest <- multiple_ttests(residuals_sv, treatment,
                group_1 = L,
                group_2 = P,
                mean_1 = mean_L,
                mean_2 = mean_P,
                values = resid,
                equal_var = TRUE)
```

Plot p-value histogram.

```{r phist}
plot_pvalue_histogram(residuals_sv_ttest, p_value, limit_max = 600)
```

```{r}
fdr_correction_with_qvalue <- function(tb, p_value_col) {
  
  # Quote
  p_value_col <- enquo(p_value_col)
  
  # fdr correction
  qvalues <- qvalue(as_vector(tb %>%
                                ungroup() %>% # To make sure to only get one column
                                dplyr::select(!!p_value_col)))
  
  return(qvalues)
}
```

FDR correction
```{r}
q_vals <- fdr_correction_with_qvalue(residuals_sv_ttest, p_value) 
```

```{r}
resids_ttest <- multiple_ttests(resids_sv, treatment,
                group_1 = L,
                group_2 = P,
                mean_1 = mean_L,
                mean_2 = mean_P,
                values = resid,
                equal_var = TRUE)
```

```{r ses_info, echo = FALSE}
sessionInfo()
```


