---
title: "R Notebook"
output: html_notebook
---

Read libraries.

```{r library, message = FALSE}
library(tidyverse)
library(modelr)
library(broom)
library(ggfortify)
library(sva)
library(qvalue)
library(ggrepel)
```

Read functions.

```{r functions, echo = FALSE}
source("functions/separate_condition_from_DEP_output.R")
source("functions/format_data_for_PCA.R")
source("functions/plot_var_exp.R")
source("functions/make_var_exp_kable.R")
source("functions/calculate_variance_explained.R")
source("functions/plot_pca.R")
source("functions/plot_multiple_PCAs.R")
source("functions/multiple_ttests.R")
source("functions/fdr_correction_with_qvalue.R")
source("functions/plot_pvalue_histogram.R")
```

# Analysis

Perform statistical analysis of imputed data with at least 5 valid values per 
group.

Read data.

```{r message = FALSE}
data <- read_tsv("GIPR_liraglutide_liver_5vv_imputed.tsv")
```

Separate conditions into separate columns.

```{r sep_col}
factored_data <- separate_condition_from_DEP_output(data, c("treatment", "gender", "series", "litter"), remove = FALSE)
```


## Linear model

Nest data.

```{r}
nested_data <- factored_data %>% 
  group_by(name) %>% 
  nest()
```

Linear model function.

```{r func}
protein_model <- function(tb) {
  lm(LFQ ~ gender, data = tb)
}
```

Perform linear regression on each protein and calculate/extract residuals.

```{r lming}
nested_model_data <- nested_data %>% 
  mutate(model = map(data, protein_model))

nested_resids_data <- nested_model_data %>% 
  mutate(resids = map2(data, model, add_residuals))

resids <- unnest(nested_resids_data, resids)
```

Calculate variance and compare with original data.

```{r var}
variance <- resids %>% 
  group_by(name) %>% 
  summarise(var_lfq = var(LFQ),
            var_resid = var(resid))

variance %>% 
  summarise(mean_lfq = mean(var_lfq),
            mean_resid = mean(var_resid))

# Test if the regression reduced the variance for all proteins
variance %>% 
  mutate(smaller = if_else(var_resid < var_lfq, 1, 0)) %>% 
  summarise(n = sum(smaller),
            all = n())

resids %>% 
  group_by(name, treatment) %>% 
  summarise(var_lfq = var(LFQ),
            var_resid = var(resid))
```

The linear model decreased the variance for every protein!

## PCA

Unite replicate (ID) with condition (grouping), and spread to intensities into
separate columns for each sample.

```{r pca_prep}
lfq_for_pca <- resids %>% 
  dplyr::select(name, replicate, condition, LFQ) %>% 
  unite(replicate, replicate, condition) %>% 
  spread(key = "replicate", value = "LFQ")

resids_for_pca <- resids %>% 
  dplyr::select(name, replicate, condition, resid) %>% 
  unite(replicate, replicate, condition) %>% 
  spread(key = "replicate", value = "resid")
```


Perform PCA for both LFQ and residuals.

```{r pca}
lfq_pca <- format_data_for_PCA(lfq_for_pca, 17, c("sample", "treatment", "gender", "series", "litter"))

resid_pca <- format_data_for_PCA(resids_for_pca, 17, c("sample", "treatment", "gender", "series", "litter"))
```

Calculate the variance explained by each principal component for original and
data processed with linear regression.

```{r var_exp}
# Calculate the variance that each PC explains
lfq_var_exp <- calculate_variance_explained(lfq_pca)
make_var_exp_kable(lfq_var_exp)

resid_var_exp <- calculate_variance_explained(resid_pca)
make_var_exp_kable(resid_var_exp)

# Plot variance explained plots
plot_var_exp(lfq_var_exp)
plot_var_exp(resid_var_exp)
```

The principal components from the linear model explains considerably more of the
variance than the LFQ intensities.


Plot the first two PCs of the PCA.

```{r plot_pca}
plot_pca(lfq_pca, "LFQ", "treatment")
plot_pca(resid_pca, "resids", "treatment")
```

# t-test

Perform multiple t-test.

```{r ttest}
residuals_ttest <- multiple_ttests(resids, treatment,
                group_1 = L,
                group_2 = P,
                mean_1 = mean_L,
                mean_2 = mean_P,
                values = resid,
                equal_var = TRUE)
```

Plot p-value histogram.

```{r phist}
plot_pvalue_histogram(residuals_ttest, p_value, limit_max = 600)
```

```{r echo = FALSE}
fdr_correction_with_qvalue <- function(tb, p_value_col) {
  
  # Quote
  p_value_col <- enquo(p_value_col)
  
  # fdr correction
  qvalues <- qvalue(as_vector(tb %>%
                                ungroup() %>% # To make sure to only get one column
                                dplyr::select(!!p_value_col)))
  
  return(qvalues)
}
```

FDR correction with `qvalue`.

```{r}
q_vals <- fdr_correction_with_qvalue(residuals_ttest, p_value) 
```
```{r}
make_qvalue_tibble <- function(tb) {
  qvalues <- tibble(p_value = tb[["pvalues"]], qvalue = tb[["qvalues"]])
  
  return(qvalues)
}
```

Join statistical test outputs with all data.

```{r}
qvalues <- make_qvalue_tibble(q_vals)
qvalues %>% arrange(qvalue)

resids_ttest <- full_join(residuals_ttest, qvalues, by = "p_value")

all_data <- full_join(factored_data, resids_ttest, by = "name")
```

# Join all data and make it ready for excel

I need to join residuals with `all_data`.

```{r}
resids
all_data <- full_join(all_data %>% unite(name, replicate, col = "join_by"),
          resids %>% unite(name, replicate, col = "join_by") %>% 
            select(join_by, resid),
          by = "join_by") %>% 
  separate(join_by, into = c("name", "replicate"), sep = "_")
```

Save `all_data`.

```{r, eval = FALSe}
write_tsv(all_data %>% 
            select(-L, -P), "data/all_data_long_unstructured.txt")
```

Make wide LFQ and residuals `.txt` files.

```{r}
# Unique data
unique_data <- all_data %>% 
  select(-label:-litter, -replicate, -LFQ, -resid, -L, -P) %>% 
  distinct()

# LFQ
lfq_wide <- all_data %>% 
  select(name, treatment:replicate, LFQ) %>%
  unite(treatment, gender, series, litter, replicate, col = "label") %>% 
  spread(key = label, value = LFQ)

lfq_wide <- full_join(lfq_wide, unique_data, by = "name")

# residuals
resids_wide <- all_data %>% 
  select(name, treatment:replicate, resid) %>%
  unite(treatment, gender, series, litter, replicate, col = "label") %>% 
  spread(key = label, value = resid)

resids_wide <- full_join(resids_wide, unique_data, by = "name")
```

Write wide `txt`files.

```{r echo = FALSE}
# LFQ
write_tsv(lfq_wide, "data/gipr_lfq_wide.txt")

# Residuals
write_tsv(resids_wide, "data/gipr_resids_wide.txt")
```

# Filter for significant proteins

```{r}
resids_sig_wide <- resids_wide %>% 
  filter(qvalue < 0.05)
```

```{r echo = FALSE}
write_tsv(resids_sig_wide, "data/gipr_sig_resids_wide.txt")
```

## Filter for protein names

```{r eval = FALSE}
source("functions/filter_for_protein_name.R")

lira_sig <- filter_for_protein_name(read_tsv("data/gipr_lira_sig.txt"))

write_tsv(lira_sig, "data/gipr_lira_sig_for_thesis.txt")

placebo_sig <- filter_for_protein_name(read_tsv("data/gipr_placebo_sig.txt"))

write_tsv(placebo_sig, "data/gipr_placebo_sig_for_thesis.txt")
```

# Volcano plot

```{r}
source("functions/prepare_data_for_volcano_plot.R")
source("functions/plo_fdr_volcano.R")
```

```{r}
resids_volcano <- prepare_data_for_volcano_plot(resids_ttest, "Placebo", "Liraglutide")

p <- plot_fdr_volcano(resids_volcano, plot_title = "", 
                 break_vector = c("Placebo", "Liraglutide"),
                 color_vector = c("#F8666D", "black", "#00BFC4"), 
                 a = 0.75)
p

p1 <- p +
  guides(color = FALSE) +
  theme(axis.title = element_text(size = 18),
        axis.text = element_text(size = 14))
p1
ggsave("GIPR_volcano_plot_without_labels.pdf", p1, device = "pdf")

p2 <- p +
  guides(color = FALSE) +
  theme(axis.title = element_text(size = 18),
        axis.text = element_text(size = 14)) +
  geom_text_repel(data = resids_volcano %>% 
                     filter(significant == "Placebo" | significant == "Liraglutide") %>% 
                     filter(log2_difference < -0.8 | log2_difference > .8), aes(log2_difference, -log10(p_value),  label = name), color = "black")
p2
ggsave("GIPR_volcano_plot_with_labels.pdf", p2, device = "pdf")
```


```{r}
resids_sig_wide %>% 
  filter(qvalue < 0.05 & log2_difference > 0) %>% 
  select(name, Protein_ID, log2_difference) %>% 
  arrange(desc(log2_difference)) %>% View
```

# Make table for GSEA

```{r, eval = FALSE}
residuals_gsea <- resids %>%
  select(name, label, resid) %>% 
  spread(key = "label", "resid")

write_tsv(residuals_gsea, "GIPR_liver_resids_table_for_gsea.txt")
```

```{r}
resids
```

# Format GSEA output

```{r, eval = FALSE}
gsea_lira <- read_tsv("data/gipr_liraglutide_gsea.txt")

gsea_lira <- gsea_lira %>% 
  separate(Name, into = c("Category", "Name"), sep = "_", extra = "merge") %>% 
  mutate(Name = tolower(Name),
         Name = str_to_title(Name),
         Name = str_replace_all(Name, "_", " "),
         Name = str_replace_all(Name, "rna", "RNA"),
         Name = str_replace_all(Name, "dna", "DNA"))

write_tsv(gsea_lira, "gsea_lira_table_for_thesis.txt")

placebo_gsea <- read_tsv("data/gipr_placebo_gsea.txt")

placebo_gsea <- placebo_gsea %>% 
  separate(Name, into = c("Category", "Name"), sep = "_", extra = "merge") %>% 
  mutate(Name = tolower(Name),
         Name = str_to_title(Name),
         Name = str_replace_all(Name, "_", " "),
         Name = str_replace_all(Name, "rna", "RNA"),
         Name = str_replace_all(Name, "dna", "DNA"))

write_tsv(placebo_gsea, "gipr_placebo_gsea_table_thesis.txt")
```



# Compare with original data.

```{r ttest}
data_ttest <- multiple_ttests(factored_data, treatment,
                group_1 = L,
                group_2 = P,
                mean_1 = mean_L,
                mean_2 = mean_P,
                values = LFQ,
                equal_var = TRUE)
```

Plot p-value histogram.

```{r phist}
plot_pvalue_histogram(data_ttest, p_value, limit_max = 600)
```



